---
title: "ML_HW4"
author: "Xiaoman Xu"
date: "2023-02-27"
output:
  pdf_document: default
  html_document: default
---

```{r}
library('MASS') ## for 'mcycle'
library('manipulate') ## for 'manipulate'
```


```{r}
y <- mcycle$accel
x <- matrix(mcycle$times, length(mcycle$times), 1)

plot(x, y, xlab="Time (ms)", ylab="Acceleration (g)")
```



# 1. Randomly split the mcycle data into training (75%) and validation (25%) subsets.
```{r}
data(mcycle)

## 75% of the sample size
smp_size <- floor(0.75 * nrow(mcycle))

## set the seed to make your partition reproducible
set.seed(123)
train_ind <- sample(seq_len(nrow(mcycle)), size = smp_size)

train <- mcycle[train_ind, ]
test <- mcycle[-train_ind, ]
```

# 2. Using the mcycle data, consider predicting the mean acceleration as a function of time. Use the Nadaraya-Watson method with the k-NN kernel function to create a series of prediction models by varying the tuning parameter over a sequence of values. (hint: the script already implements this)

```{r}
kernel_epanechnikov <- function(x, x0, lambda=1) {
  d <- function(t)
    ifelse(t <= 1, 3/4*(1-t^2), 0)
  z <- t(t(x) - x0)
  d(sqrt(rowSums(z*z))/lambda)
}

## k-NN kernel function
## x  - n x p matrix of training inputs
## x0 - 1 x p input where to make prediction
## k  - number of nearest neighbors
kernel_k_nearest_neighbors <- function(x, x0, k=1) {
  ## compute distance betwen each x and x0
  z <- t(t(x) - x0)
  d <- sqrt(rowSums(z*z))

  ## initialize kernel weights to zero
  w <- rep(0, length(d))
  
  ## set weight to 1 for k nearest neighbors
  w[order(d)[1:k]] <- 1
  
  return(w)
}

## Make predictions using the NW method
## y  - n x 1 vector of training outputs
## x  - n x p matrix of training inputs
## x0 - m x p matrix where to make predictions
## kern  - kernel function to use
## ... - arguments to pass to kernel function
nadaraya_watson <- function(y, x, x0, kern, ...) {
  k <- t(apply(x0, 1, function(x0_) {
    k_ <- kern(x, x0_, ...)
    k_/sum(k_)
  }))
  yhat <- drop(k %*% y)
  attr(yhat, 'k') <- k
  return(yhat)
}

## Helper function to view kernel (smoother) matrix
matrix_image <- function(x) {
  rot <- function(x) t(apply(x, 2, rev))
  cls <- rev(gray.colors(20, end=1))
  image(rot(x), col=cls, axes=FALSE)
  xlb <- pretty(1:ncol(x))
  xat <- (xlb-0.5)/ncol(x)
  ylb <- pretty(1:nrow(x))
  yat <- (ylb-0.5)/nrow(x)
  axis(3, at=xat, labels=xlb)
  axis(2, at=yat, labels=ylb)
  mtext('Rows', 2, 3)
  mtext('Columns', 3, 3)
}
```


# 3. With the squared-error loss function, compute and plot the training error, AIC, BIC, and validation error (using the validation data) as functions of the tuning parameter.

```{r}

## Compute effective df using NW method
## y  - n x 1 vector of training outputs
## x  - n x p matrix of training inputs
## kern  - kernel function to use
## ... - arguments to pass to kernel function
effective_df <- function(y, x, kern, ...) {
  y_hat <- nadaraya_watson(y, x, x,
    kern=kern, ...)
  sum(diag(attr(y_hat, 'k')))
}

# loss function
## y    - train/test y
## yhat - predictions at train/test x
loss_squared_error <- function(y, yhat)
  (y - yhat)^2

## test/train error
## y    - train/test y
## yhat - predictions at train/test x
## loss - loss function
error <- function(y, yhat, loss=loss_squared_error)
  mean(loss(y, yhat))

## AIC
## y    - training y
## yhat - predictions at training x
## d    - effective degrees of freedom
aic <- function(y, yhat, d)
  error(y, yhat) + 2/length(y)*d

## BIC
## y    - training y
## yhat - predictions at training x
## d    - effective degrees of freedom
bic <- function(y, yhat, d)
  error(y, yhat) + log(length(y))/length(y)*d



## make predictions using NW method at training inputs
y_hat <- nadaraya_watson(y, x, x,
  kernel_epanechnikov, lambda=5)

## view kernel (smoother) matrix
matrix_image(attr(y_hat, 'k'))

## compute effective degrees of freedom
edf <- effective_df(y, x, kernel_epanechnikov, lambda=5)
aic(y, y_hat, edf)
bic(y, y_hat, edf)

## create a grid of inputs 
x_plot <- matrix(seq(min(x),max(x),length.out=100),100,1)

## make predictions using NW method at each of grid points
y_hat_plot <- nadaraya_watson(y, x, x_plot,
  kernel_epanechnikov, lambda=1)

## plot predictions
plot(x, y, xlab="Time (ms)", ylab="Acceleration (g)")
lines(x_plot, y_hat_plot, col="#882255", lwd=2) 

```


# 4. For each value of the tuning parameter, Perform 5-fold cross-validation using the combined training and validation data. This results in 5 estimates of test error per tuning parameter value.

```{r}
library('manipulate')
library('splines') ## 'ns'
#install.packages('caret')
library('caret') ## 'knnreg' and 'createFolds'
```


```{r}
## 5-fold cross-validation of knnreg model
## create five folds
set.seed(1985)
inc_flds  <- createFolds(mcycle$accel, k=5)
print(inc_flds)
sapply(inc_flds, length)  ## not all the same length

cvknnreg <- function(kNN = 10, flds=inc_flds) {
  cverr <- rep(NA, length(flds))
  for(tst_idx in 1:length(flds)) { ## for each fold
    
    ## get training and testing data
    inc_trn <- inc[-flds[[tst_idx]],]
    inc_tst <- inc[ flds[[tst_idx]],]
    
    ## fit kNN model to training data
    knn_fit <- knnreg(mcycle$accel ~ mcycle$times,
                      k=kNN, data=inc_trn)
    
    ## compute test error on testing data
    pre_tst <- predict(knn_fit, inc_tst)
    cverr[tst_idx] <- mean((mcycle$accel - pre_tst)^2)
  }
  return(cverr)
}

## Compute 5-fold CV for kNN = 1:20
cverrs <- sapply(1:20, cvknnreg)
print(cverrs) ## rows are k-folds (1:5), cols are kNN (1:20)
cverrs_mean <- apply(cverrs, 2, mean)
cverrs_sd   <- apply(cverrs, 2, sd)


```

# 5.Plot the CV-estimated test error (average of the five estimates from each fold) as a function of the tuning parameter. Add vertical line segments to the figure (using the segments function in R) that represent one “standard error” of the CV-estimated test error (standard deviation of the five estimates from each fold).
```{r}
## Plot the results of 5-fold CV for kNN = 1:20
plot(x=1:20, y=cverrs_mean, 
     ylim=range(cverrs),
     xlab="'k' in kNN", ylab="CV Estimate of Test Error")
segments(x0=1:20, x1=1:20,
         y0=cverrs_mean-cverrs_sd,
         y1=cverrs_mean+cverrs_sd)
best_idx <- which.min(cverrs_mean)
points(x=best_idx, y=cverrs_mean[best_idx], pch=20)
abline(h=cverrs_mean[best_idx] + cverrs_sd[best_idx], lty=3)
```

# 6. Interpret the resulting figures and select a suitable value for the tuning parameter.
According to the plot, we can see that when k = 1, it will generate the smallest Test Error.
 


